Territoriality and the organization of technology during the Last Glacial Maximum in southwestern Europe

### 2019-10-31 analysis/paper/paper.Rmd
After revision

# Abstract
Climate changes that occurred during the Last Glacial Maximum (LGM) had significant consequences in human eco-dynamics across Europe. Among the most striking impacts are the demographic contraction of modern humans into southern refugia and the potential formation of a population bottleneck. In Iberia and southern France transformations also included the occurrence of significant technological changes, mostly marked by the emergence of a diverse set of bifacially-shaped stone projectiles. The rapid dissemination of bifacial technologies and the geographical circumscription of specific projectile morphologies within these regions have been regarded as evidence for: (1) the existence of a system of long-distance exchange and social alliance networks; (2) the organization of human groups into cultural facies with well-defined stylistic territorial boundaries. However, the degree and modes in which cultural transmission have occurred within these territories, and how it may have influenced other domains of the adaptive systems, remains largely unknown. Using southern Iberia as a case-study, this paper presents the first quantitative approach to the organization of lithic technology and its relationship to hunter-gatherers' territorial organization during the LGM. Similarities and dissimilarities in the presence of metric and morphological data describing lithic technologies are used as a proxy to explore modes and degrees of cultural transmission. Statistical results show that similarities in technological options are dependent on the chronology and geographical distance between sites and corroborate previous arguments for the organization of LGM settlement in Southern Iberia into discrete eco-cultural facies.

@@

Climatic changes occurred during this period had substantial repercussions in the geography, technology, and social behavior of human populations in Europe. Among the most drastic impacts are: (1) the contraction of the human range to southern regions of Europe that served as refugia, such as the Iberian and the Italian peninsulas [@banks_human_2008; @bocquet-appel_population_2000; @bradtmoller_repeated_2012; @burke_exploring_2014; @derrico_neandertal_2003; @jochim_late_1987; @straus_riera_1986; @straus_southwestern_1991]; (2) the consequent genesis of a bottleneck scenario in human genetic diversity [@barbujani_evidence_1998; @keinan_measurement_2007; @torroni_signal_2001]; (3) and, in southwestern Europe, the advent of a set of new technologies for the manufacture of lithic projectiles and knives using bifacial retouch.

@@

```{r lgmmap, echo=FALSE, fig.cap = "Top - Geographical distribution of the cultural facies (A- Cantabrian, B- Portuguese, C- Mediterranean, D-Catalan) defined for the LGM in Iberia. Bottom -  location of LGM sites in the southern part of the Peninsula, sites used in this study are marked with a star."}

@@

LGM settlement of southern Iberia is marked by the existence of more or less dense clusters of sites, that are relatively isolated from one another by geographical hiatus where the complete absence or the presence of very few sites might testify human passage but not likely intensive occupation [@aura_cueva_2009; @cascalheira_vale_2012]. Caves and rock-shelters are better represented than open-air locations, as are multi-component sites over single horizon ones [@cascalheira_chronological_2015; @cascalheira_influencia_2013]. There is evidence for the existence of different types of sites, including long-term and seasonal residential camps, shorter specialized occupations, lithic cache sites, and possible aggregation sites [sensu @conkey_identification_1980]. The number of occupied sites is significantly higher than during the early Upper Paleolithic and, although they tend to be in coastal or near-coastal settings, inland camps are also represented, mainly as single-occupation localities [@schmidt_rapid_2012; @straus_upper_2000].

@@

The diffusion of specific concepts across southern Iberia, as rapidly as it seems to have occurred during the LGM, may have resulted from an array of very diverse contact behaviors. Ethnoarchaeological data [@wiessner_reconsidering_1984; @wiessner_risk_1982; @wiessner_style_1983] has demonstrated that the process of social interaction between hunter-gatherer communities can vary between episodes in which people have little contact to each other’s domestic space (with gatherings occurring at the edge of their respective ranges), to events in which marriage partners are exchanged, and previously foreign individuals become fixtures in each other’s residential lives [@tostevin_seeing_2013; @tostevin_social_2007]. Contacts occurring anywhere between these two extremes of the contact scale result in dissimilar access to information, and to different types of information.

@@

Such kind of analysis requires, however, a theoretical and methodological framework that would allow the integration of both social interactions and ecological background as shapers of stone knapping patterns across space and time. While there are many ethnoarchaeological and theory-based studies addressing the meaning of morphological variations of lithic tools as indicators of style and territorial differentiation, very few attempts have been made to use technological processes as proxies for cultural transmission processes. The middle-range theory presented by Tostevin [-@tostevin_seeing_2013; -@tostevin_social_2007], for studying the material evidence of social contact between Neanderthals and Anatomically Modern Humans using lithic technology, is perhaps one of the most compelling and sophisticated approaches. Based on a body of literature on artefact style [@jochim_building_1995; @jochim_unified_1995; @lechtman_style_1975; @sackett_style_1990; @wiessner_reconsidering_1984; @wiessner_seeking_1997], optimization studies [@bird_behavioral_2006; @boyd_culture_1985; @broughton_evolutionary_1999; @kuhn_evolutionary_2004] and experimental knapping [@dibble_new_1981; @pelcin_effect_1997; @pelcin_effect_1997-1; @pelcin_formation_1997], Tostevin’s model seeks to identify whether variation in lithic material culture style reflects one or more social processes, once other sources of variability have been dismissed.

@@

Ethnography has shown that technological gestures are not just individual action sequences with clearly defined goals, but suites of distributed behaviors reflecting complex systems that have been learned and assimilated as bodily techniques. In this context, flintknapping processes are considered “technological performances which can only be effectively taught and learned by observation of the body movements, the handling of core and percussor, and the flake by flake strategizing on how to exploit core volumes” [@tostevin_social_2007, pp. 344]. Lithic reduction is a directional and irreversible process characterized by a sequence of mechanical relationships. While these relationships are influenced by multiple contextual, aspects, such as the quality of the available raw material, identifying different exploitation patterns within these mechanical relationships are thought to reflect a purposeful choice of particular modes of technological manipulation present in an assemblage out of an equifinal pool of alternative possibilities - what Sackett [-@sackett_approaches_1982; -@sackett_style_1986; -@sackett_style_1990] have called isochrestic variation .

@@

knitr::kable(expectations_table, caption = "Model expectations for different degrees of cultural transmission during the LGM in southern Iberia based on technological analysis of lithic attributes.")

@@

Following the theoretical background of Tostevin [-@tostevin_seeing_2013; -@tostevin_social_2007], Scerri [-@scerriAterianItsPlace2013] and Scerri et al. [-@scerri_unexpected_2014], in this study variables were organized into comparable heuristic domains (i.e. Platform Maintenance, Core Exploitation, Dorsal Convexities Management, Elongated blanks Production, Core Morphology, and Core Use) of the technological sequences, to achieve a comprehensive structure of similarity and dissimilarity between assemblages. The main goal of using these domains was to isolate different sources of variability by focusing on specific sets of interrelated variables that correspond to interdependent knapping actions. Breaking down the reduction sequences into comparable clusters makes possible to account for possible biased samples and “broken” sequences, not assuming that the knapping process was restricted to a specific timeframe, or the work of a single individual [@scerri_unexpected_2014].

@@

A multivariate statistical approach was used to explore the interaction of variables within each domain. Since most of the attributes available were categorical, a Multiple Correspondence Analysis (MCA) was used to explore the orthogonal dimensions of variability in the data. MCA is well suited to lithic analysis because it allows the isolation of the different sources of variability contributing to the character of each assemblage under study. When used, metric variables were pooled into categories using a K-means clustering algorithm. To reduce the amount of ‘noise’ in the overall variability, all categorical attributes whose representation was less than 5% within each variable were also lumped together into a category named ‘Other’, before any of the analysis were performed.

@@

total_blanks <- left_join(cores_prop, blanks_prop) %>% 
  left_join(., cortex_ratio) %>% 
  mutate(`Blank to Core ratio` = Blanks / Cores) %>% 
  mutate(`Total artifacts` = Cores + Blanks)

@@

```{r histogram, echo = FALSE, fig.cap = "Width histograms for the elongated products from each context included in this study."}

@@

ggplot(bladelets, aes(Width)) +
  geom_histogram(aes(y = ..density..), color = "grey", fill = "grey") +
  geom_density() +
  facet_wrap(~Context, ncol = 2) +
  scale_y_continuous(labels = scales::percent)

@@

```{r intensity, echo = FALSE, fig.cap = "Density plots comparing the lengths of dominant blank scars on cores (yellow), blanks (blue) and fully cortical blanks (magenta). Following Henry [-@henry_correlations_1989] the plots show the relative flaking intensity between the assemblages. For interpretation of the references to colour in this figure legend, the reader is referred to the web version of this article."}

@@

ggplot() + 
  geom_density(data = core_length, aes(x=ScarLength), fill = "#E69F00", color = "black", alpha = 0.7) +
  geom_density(data = blank_length, aes(x=Length), fill = "#56B4E9", color = "black", alpha = 0.7) +
  geom_density(data = primary_blanks_length, aes(x=Length), fill = "#ff2d2d", color = "black", alpha = 0.5) +
  facet_wrap(~Context, ncol = 2) +
  labs(x = "Length (mm)")

@@

```{r MCAplatformmaintenance, echo=FALSE, fig.cap = "Multiple Correspondence Analysis plot for Platform Maintenance. For interpretation of the references to colour in this figure legend, the reader is referred to the web version of this article."}

@@

fviz_mca_var(mca_res_platform_maintenance, labelsize = 3, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray(), title = "") +
             theme(axis.text.x = element_text(size = 14),
            axis.text.y = element_text(size = 14), text = element_text(size = 18))

@@

```{r ANOVAplatformmaintenance, echo=FALSE, fig.cap = "Platform Maintenance multi-comparison bootstrap ANOVA. Error bars represent 95% confidence interval from the mean. Contexts sharing the same set of letters do not present statistically significant difference."}

@@

```{r MCAcoreexploitation, echo=FALSE, fig.cap = "Multiple Correspondence Analysis plot for Core Exploitation. For interpretation of the references to colour in this figure legend, the reader is referred to the web version of this article."}

@@

fviz_mca_var(mca_res_core_exploitation, labelsize = 3, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray(), title = "") +
  theme(axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14), text = element_text(size = 18))

@@

```{r ANOVAcoreexploitation, echo=FALSE, fig.cap = "Core Exploitation multi-comparison bootstrap ANOVA. Error bars represent 95% confidence interval from the mean. Contexts sharing the same set of letters do not present statistically significant difference."}

@@

The influence of flaking surface convexities in the overall morpho-metrics of a blank has been demonstrated by sereval authors [see e.g., @macgregor_abrupt_2005; @pelcin_controlled_1996; @tostevin_social_2007]. In short, the ridge pattern of the dorsal surface of flakes determines the path of the resulting flake because the ridges guide the force of the percussor blow. Changes in the shape of the debitage surface morphology affect the dimensions of the resulting blanks by changing the distribution of mass over the said blank. In this analytical domain, a set of variables related to shape, including Elongation and Flattening ratios, Cross-section, Edge shape, and Profile are examined for all blanks. The aim of this approach is to assess patterns of blank morphological variation of each context under study, to evaluate possible differences/similiarities on how knappers controlled surface convexities during core exploitation.

@@

```{r MCAconvexity, echo=FALSE, fig.cap = "Multiple Correspondence Analysis plot for Dorsal Surface Convexity. For interpretation of the references to colour in this figure legend, the reader is referred to the web version of this article."}

@@

fviz_mca_var(mca_res_dorsal_convexity, labelsize = 3, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray(), title = "") +
  theme(axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14), text = element_text(size = 18))

@@

```{r ANOVAconvexity, echo=FALSE, fig.cap = "Dorsal Convexity multi-comparison bootstrap ANOVA. Error bars represent 95% confidence interval from the mean. Contexts sharing the same set of letters do not present statistically significant difference."}

@@

```{r MCAelongated, echo=FALSE, fig.cap = "Multiple Correspondence Analysis plot for Elongated Products. For interpretation of the references to colour in this figure legend, the reader is referred to the web version of this article."}

@@

fviz_mca_var(mca_res_elongated, labelsize = 3, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray(), title = "") +
  theme(axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14), text = element_text(size = 18))

@@

```{r ANOVAelongated, echo=FALSE, fig.cap = "Elongated Products multi-comparison bootstrap ANOVA. Error bars represent 95% confidence interval from the mean. Contexts sharing the same set of letters do not present statistically significant difference."}

@@

```{r MCAcoremorphology, echo=FALSE, fig.cap = "Multiple Correspondence Analysis plot for Core Morphology. For interpretation of the references to colour in this figure legend, the reader is referred to the web version of this article."}

@@

fviz_mca_var(mca_res_cores, labelsize = 3, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray(), title = "") +
  theme(axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14), text = element_text(size = 18))

@@

```{r ANOVAcoremorphology, echo=FALSE, fig.cap = "Core Morphology multi-comparison bootstrap ANOVA. Error bars represent 95% confidence interval from the mean. Contexts sharing the same set of letters do not present statistically significant difference."}

@@

```{r MCAcoreuse, echo=FALSE, fig.cap = "Multiple Correspondence Analysis plot for Core Use. For interpretation of the references to colour in this figure legend, the reader is referred to the web version of this article."}

@@

fviz_mca_var(mca_res_core_use, labelsize = 3, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray(), title = "") +
  theme(axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14), text = element_text(size = 18))

@@

```{r ANOVAcoreuse, echo=FALSE, fig.cap = "Core Use multi-comparison bootstrap ANOVA. Error bars represent 95% confidence interval from the mean. Contexts sharing the same set of letters do not present statistically significant difference."}

@@

```{r ANOVAsimilarity, echo = FALSE, fig.cap= "Heat map summarizing the number of non-significantly different bootstrapped ANOVA tests between each pair of assemblages. For interpretation of the references to colour in this figure legend, the reader is referred to the web version of this article."}

@@

@@

 coord_fixed()

@@

In the first instance, chronological change in the technological options of lithic production is attested by significant dissimilarities between some contexts within multi-layered sites. This pattern is mostly visible in the particularities of platform maintenance and core exploitation domains. However, this seems to have only occurred at AMB and PAP. At AMB, two general technological phases can be distinguished, with results differentiating between the oldest contexts (AMB VI and AMB IV) and the more recent level (AMB II). At PAP, the separation also occurs in the transition between the two oldest levels (PAP 5’25-6’25 and PAP 4’75-5’25) and the most recent one (PAP 4’00-4’75). These disruptions are, at both sites, coincident with the transition from the traditionally-defined Upper Solutrean to Solutreo-Gravettian (or Evolved Upper Solutrean) phases, which is thought to have occurred, at these sites, at around c. 20.5 ka cal BP [@munoz_ibanez_puntas_2000; @fullola_industrias_1979; @ripoll_lopez_cueva_1988]. Thus, previous differences detected between those phases, based on typological grounds and pointing to a decline in the use of flat-retouched points and the increasing presence of backed shouldered points [@aura_solutrenses_1989; @tiffagom_pierre_2006], seem to be corroborated by significant changes in the technological substrate. The most important changes regarding this transition comprise an increasing use of reduction sequences oriented to the production of highly elongated blanks, using opposed bidirectional methods on prismatically-organized cores. The absence of this tendency at VB [see e.g., @cascalheira_chronological_2015; @cascalheira_influencia_2013] might result from a lack of effective occupation at the site during the later parts of the LGM [which is rather probable given that the latest radiocarbon date for the solutrean occupations at the site is c. 21 ka cal BP - @cascalheira_chronological_2015; @cascalheira_vale_2012], but also by technological drift occuring during this period across Iberia, as suggested by the fact that a Solutreo-Gravettian phase was, so far, not clearly recognized in the Atlantic facade of Iberia.

>>>>

# Abstract
Climate changes that occurred during the Last Glacial Maximum (LGM) had significant consequences in human eco-dynamics across Europe. Among the most striking impacts are the demographic contraction of modern humans into southern refugia and the potential formation of a population bottleneck. In Iberia and southern France transformations also included the occurrence of significant technological changes, mostly marked by the emergence of a diverse set of bifacially-shaped stone projectiles. The rapid dissemination of bifacial technologies and the geographical circumscription of specific projectile morphologies within these regions have been regarded as evidence for: (1) the existence of a system of long-distance exchange and social alliance networks; (2) the organization of human groups into cultural facies with well-defined stylistic territorial boundaries. However, the degree and modes in which cultural transmission have occurred within these territories, and how it may have influenced other domains of the adaptive systems, remains largely unknown. Using southern Iberia as a case-study, this paper presents the first quantitative approach to the organization of lithic technology and its relationship to hunter-gatherers' territorial organization during the LGM. Similarities and dissimilarities in the presence of morphological and metrics data describing lithic technologies are used as a proxy to explore modes and degrees of cultural transmission. Statistical results show that similarities in technological options are dependent on the chronology and geographical distance between sites and corroborate previous arguments for the organization of LGM settlement in Southern Iberia into discrete eco-cultural facies.

@@

Climatic changes occurred during this period had substantial repercussions in the geography, technology, and social behavior of human populations in Europe. Among the most drastic impacts are: (1) the contraction of the human range to southern regions of Europe that served as refugia, such as the Iberian and the Italian peninsulas [@banksHumanEcologicalNiches2008; @bocquet-appel_population_2000; @bradtmollerRepeatedReplacementModel2012; @burke_exploring_2014; @derricoNeandertalExtinctionMillennial2003; @jochimLatePleistoceneRefugia1987; @straus_riera_1986; @straus_southwestern_1991]; (2) the consequent genesis of a bottleneck scenario in human genetic diversity [@barbujani_evidence_1998; @keinan_measurement_2007; @torroni_signal_2001]; (3) and, in southwestern Europe, the advent of a set of new technologies for the manufacture of lithic projectiles and knives using bifacial retouch.

@@

```{r lgmmap, echo=FALSE, fig.cap = "Top - Geographical distribution of the cultural facies (A- Cantabrian, B- Portuguese, C- Mediterranean, D-Catalan) defined for the LGM in Iberia. Bottom -  location of LGM sites in the southern part of the Peninsula, sites used in this study are marked with a star. Topographic data from the U.S. Geological, Survey Shuttle Radar Topography Mission (https://earthexplorer.usgs.gov/)."}

@@

LGM settlement of southern Iberia is marked by the existence of more or less dense clusters of sites, that are relatively isolated from one another by geographical hiatus where the complete absence or the presence of very few sites might testify human passage but not likely intensive occupation [@aura_cueva_2009; @cascalheira_vale_2012]. Caves and rock-shelters are better represented than open-air locations, as are multi-component sites over single horizon ones [@cascalheira_chronological_2015; @cascalheira_influencia_2013]. There is evidence for the existence of different types of sites, including long-term and seasonal residential camps, shorter specialized occupations, lithic cache sites, and possible aggregation sites [sensu @conkey_identification_1980]. The number of occupied sites is significantly higher than during the early Upper Paleolithic and, although they tend to be in coastal or near-coastal settings, inland camps are also represented, mainly as single-occupation localities [@schmidtRapidClimateChange2012; @straus_upper_2000].

@@

The diffusion of specific concepts across southern Iberia, as rapidly as it seems to have occurred during the LGM, may have resulted from an array of very diverse contact behaviors. Ethnoarchaeological data [@wiessner_reconsidering_1984; @wiessner_risk_1982; @wiessner_style_1983] has demonstrated that the process of social interaction between hunter-gatherer communities can vary between episodes in which people have little contact to each other’s domestic space (with gatherings occurring at the edge of their respective ranges), to events in which marriage partners are exchanged, and previously foreign individuals become fixtures in each other’s residential lives [@tostevin_seeing_2013; @tostevinSocialIntimacyArtefact2007]. Contacts occurring anywhere between these two extremes of the contact scale result in dissimilar access to information, and to different types of information.

@@

Such kind of analysis requires, however, a theoretical and methodological framework that would allow the integration of both social interactions and ecological background as shapers of stone knapping patterns across space and time. While there are many ethnoarchaeological and theory-based studies addressing the meaning of morphological variations of lithic tools as indicators of style and territorial differentiation, very few attempts have been made to use technological processes as proxies for cultural transmission processes. The middle-range theory presented by Tostevin [-@tostevin_seeing_2013; -@tostevinSocialIntimacyArtefact2007], for studying the material evidence of social contact between Neanderthals and Anatomically Modern Humans using lithic technology, is perhaps one of the most compelling and sophisticated approaches. Based on a body of literature on artefact style [@jochim_building_1995; @jochim_unified_1995; @lechtman_style_1975; @sackett_style_1990; @wiessner_reconsidering_1984; @wiessner_seeking_1997], optimization studies [@bird_behavioral_2006; @boyd_culture_1985; @broughton_evolutionary_1999; @kuhn_evolutionary_2004] and experimental knapping [@dibble_new_1981; @pelcin_effect_1997; @pelcin_effect_1997-1; @pelcin_formation_1997], Tostevin’s model seeks to identify whether variation in lithic material culture style reflects one or more social processes, once other sources of variability have been dismissed.

@@

Ethnography has shown that technological gestures are not just individual action sequences with clearly defined goals, but suites of distributed behaviors reflecting complex systems that have been learned and assimilated as bodily techniques. In this context, flintknapping processes are considered “technological performances which can only be effectively taught and learned by observation of the body movements, the handling of core and percussor, and the flake by flake strategizing on how to exploit core volumes” [@tostevinSocialIntimacyArtefact2007, pp. 344]. Lithic reduction is a directional and irreversible process characterized by a sequence of mechanical relationships. While these relationships are influenced by multiple contextual, aspects, such as the quality of the available raw material, identifying different exploitation patterns within these mechanical relationships are thought to reflect a purposeful choice of particular modes of technological manipulation present in an assemblage out of an equifinal pool of alternative possibilities - what Sackett [-@sackett_approaches_1982; -@sackett_style_1986; -@sackett_style_1990] have called isochrestic variation .

@@

knitr::kable(expectations_table, caption = "Model expectations for different degrees of cultural transmission during the LGM in southern Iberia, based on technological analysis of lithic attributes.")

@@

Following the theoretical background of Tostevin [-@tostevin_seeing_2013; -@tostevinSocialIntimacyArtefact2007], Scerri [-@scerriAterianItsPlace2013] and Scerri et al. [-@scerri_unexpected_2014], in this study variables were organized into comparable heuristic domains (i.e. Platform Maintenance, Core Exploitation, Dorsal Convexities Management, Elongated blanks Production, Core Morphology, and Core Use) of the technological sequences, to achieve a comprehensive structure of similarity and dissimilarity between assemblages. The main goal of using these domains was to isolate different sources of variability by focusing on specific sets of interrelated variables that correspond to interdependent knapping actions. Breaking down the reduction sequences into comparable clusters makes possible to account for possible biased samples and “broken” sequences, not assuming that the knapping process was restricted to a specific timeframe, or the work of a single individual [@scerri_unexpected_2014].

@@

A multivariate statistical approach was used to explore the interaction of variables within each domain. Since most of the attributes available were categorical, a Multiple Correspondence Analysis (MCA) was used to explore the orthogonal dimensions of variability in the data. MCA is well suited to lithic analysis because it allows the isolation of the different sources of variability influencing the character of each assemblage under study. When used, metric variables were pooled into categories using a K-means clustering algorithm. To reduce the amount of ‘noise’ in the overall variability, all categorical attributes whose representation was less than 5% within each variable were also lumped together into a category named ‘Other’, before any of the analysis were performed.

@@

total_blanks <- left_join(cores_prop, blanks_prop) %>% 
  left_join(., cortex_ratio) %>% 
  mutate(`Blanks / Cores ratio` = Blanks / Cores) %>% 
  mutate(`Total artifacts` = Cores + Blanks)

@@

```{r histogram, echo = FALSE, fig.cap = "Width histograms for the elongated products of each context included in this study."}

@@

ggplot(bladelets, aes(Width)) +
  geom_histogram(aes(y = ..density..), color = "grey", fill = "grey") +
  geom_density() +
  facet_wrap(~Context, ncol = 2) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.title.y = element_blank()) +
  labs(x = "Width (mm)")

@@

```{r intensity, echo = FALSE, fig.cap = "Density plots comparing the lengths of dominant blank scars on cores, blanks and fully cortical (primary) blanks. Following Henry [-@henry_correlations_1989] the plots show the relative flaking intensity for each assemblage."}

@@

cols <- c("Cores" = "#E69F00", "Blanks" = "#56B4E9", "Primary Blanks" = "#ff2d2d")

ggplot() + 
  geom_density(data = core_length, aes(x=ScarLength, fill = "Cores"), color = "black", alpha = 0.7) +
  geom_density(data = blank_length, aes(x=Length, fill = "Blanks"), color = "black", alpha = 0.7) +
  geom_density(data = primary_blanks_length, aes(x=Length, fill = "Primary Blanks"), color = "black", alpha = 0.5) +
  scale_fill_manual(name=NULL,values=cols)  +
  facet_wrap(~Context, ncol = 2) +
  labs(x = "Length (mm)")

@@

```{r MCAplatformmaintenance, echo=FALSE, fig.cap = "Multiple Correspondence Analysis plot (first two dimensions) for Platform Maintenance. The color ramp represents the contribution (in %) of the variable categories to the definition of the dimensions."}

@@

fviz_mca_var(mca_res_platform_maintenance, labelsize = 5, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray(), title = "") +
             theme(axis.text.x = element_text(size = 14),
            axis.text.y = element_text(size = 14), text = element_text(size = 18))

@@

```{r ANOVAplatformmaintenance, echo=FALSE, fig.cap = "Platform Maintenance multi-comparison bootstrap ANOVA. Error bars represent 95% confidence interval from the mean. Contexts sharing the same set of letters do not present statistically significant difference (p-value ≥ 0.05)."}

@@

```{r MCAcoreexploitation, echo=FALSE, fig.cap = "Multiple Correspondence Analysis plot (first two dimensions) for Core Exploitation. The color ramp represents the contribution (in %) of the variable categories to the definition of the dimensions."}

@@

fviz_mca_var(mca_res_core_exploitation, labelsize = 5, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray(), title = "") +
  theme(axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14), text = element_text(size = 18))

@@

```{r ANOVAcoreexploitation, echo=FALSE, fig.cap = "Core Exploitation multi-comparison bootstrap ANOVA. Error bars represent 95% confidence interval from the mean. Contexts sharing the same set of letters do not present statistically significant difference (p-value ≥ 0.05)."}

@@

The influence of flaking surface convexities in the overall morpho-metrics of a blank has been demonstrated by sereval authors [see e.g., @macgregor_abrupt_2005; @pelcin_controlled_1996; @tostevinSocialIntimacyArtefact2007]. In short, the ridge pattern of the dorsal surface of flakes determines the path of the resulting flake because the ridges guide the force of the percussor blow. Changes in the shape of the debitage surface morphology affect the dimensions of the resulting blanks by changing the distribution of mass over the said blank. In this analytical domain, a set of variables related to shape, including Elongation and Flattening ratios, Cross-section, Edge shape, and Profile are examined for all blanks. The aim of this approach is to assess patterns of blank morphological variation of each context under study, to evaluate possible differences/similiarities on how knappers controlled surface convexities during core exploitation.

@@

```{r MCAconvexity, echo=FALSE, fig.cap = "Multiple Correspondence Analysis plot (first two dimensions) for Dorsal Surface Convexity. The color ramp represents the contribution (in %) of the variable categories to the definition of the dimensions."}

@@

fviz_mca_var(mca_res_dorsal_convexity, labelsize = 5, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray(), title = "") +
  theme(axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14), text = element_text(size = 18))

@@

```{r ANOVAconvexity, echo=FALSE, fig.cap = "Dorsal Convexity multi-comparison bootstrap ANOVA. Error bars represent 95% confidence interval from the mean. Contexts sharing the same set of letters do not present statistically significant difference (p-value ≥ 0.05)."}

@@

```{r MCAelongated, echo=FALSE, fig.cap = "Multiple Correspondence Analysis plot (first two dimensions) for Elongated Products. The color ramp represents the contribution (in %) of the variable categories to the definition of the dimensions"}

@@

fviz_mca_var(mca_res_elongated, labelsize = 5, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray(), title = "") +
  theme(axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14), text = element_text(size = 18))

@@

```{r ANOVAelongated, echo=FALSE, fig.cap = "Elongated Products multi-comparison bootstrap ANOVA. Error bars represent 95% confidence interval from the mean. Contexts sharing the same set of letters do not present statistically significant difference (p-value ≥ 0.05)."}

@@

```{r MCAcoremorphology, echo=FALSE, fig.cap = "Multiple Correspondence Analysis plot (first two dimensions) for Core Morphology. The color ramp represents the contribution (in %) of the variable categories to the definition of the dimensions"}

@@

fviz_mca_var(mca_res_cores, labelsize = 5, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray(), title = "") +
  theme(axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14), text = element_text(size = 18))

@@

```{r ANOVAcoremorphology, echo=FALSE, fig.cap = "Core Morphology multi-comparison bootstrap ANOVA. Error bars represent 95% confidence interval from the mean. Contexts sharing the same set of letters do not present statistically significant difference (p-value ≥ 0.05)."}

@@

```{r MCAcoreuse, echo=FALSE, fig.cap = "Multiple Correspondence Analysis plot (first two dimensions) for Core Use. The color ramp represents the contribution (in %) of the variable categories to the definition of the dimensions"}

@@

fviz_mca_var(mca_res_core_use, labelsize = 5, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray(), title = "") +
  theme(axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14), text = element_text(size = 18))

@@

```{r ANOVAcoreuse, echo=FALSE, fig.cap = "Core Use multi-comparison bootstrap ANOVA. Error bars represent 95% confidence interval from the mean. Contexts sharing the same set of letters do not present statistically significant difference (p-value ≥ 0.05)."}

@@

```{r ANOVAsimilarity, echo = FALSE, fig.cap= "Heat map summarizing the number of non-significantly different bootstrapped ANOVA tests between each pair of assemblages. Letters inside each square indicate the heuristic domains responsible for the similarities between assemblages: A - Platform Maintenance; B - Direction of Core Exploitation; C - Dorsal Surface Convexity; D - Elongated blanks techno-morphology; E - Core Morphology; F - Core Use. For more information on each of the contexts please refer to Figure 1 and Table 2."}

@@

tests <- c("   B C\nD E F",  "   B C\nD E F", "A B C\nD E F", "A B C\nD E F", "        \nD E   ", "   B   \nD E F", "   B   \n   E F", "         \n   E F", "        \nD E F", "A B C\nD E F", "A     C\nD E F", "A B   \nD E F", "       C\nD E F", "  B   \n  E F", "   B   \n   E F", "A     \n        ", "A       \n   E F",  "       C\nD E F", "A B C\nD E F", "        \nD E F", "A B   \nD E F", "A B   \n  E F", "A      \n      F", "A      \nD E F", "A      \nD E F", "        \nD E F", "  B   \nD E F", "A      \n      F", "A      \n      F", "        \n   E F", "         \nD E F", "A     C\nD E F", "A B   \n   E F", "A      \n       F", "A B   \nD E F", "A      \nD E F", "   B C\n   E F", "A B C\n  E F", "   B C\nD E F","A    C\n   E F", "       C\nD E F", "   B C\nD E F", "A     C\nD E F", "A    C\n  E F", "   B C\nD     F")

@@

 coord_fixed() +
 geom_text(aes(label = tests), size = 3)

@@

In the first instance, chronological change in the technological options of lithic production is attested by significant dissimilarities between some contexts within multi-layered sites. This pattern is mostly visible in the particularities of platform maintenance and core exploitation domains. However, this seems to have only occurred at AMB and PAP. At AMB, two general technological phases can be distinguished, with results differentiating between the oldest contexts (AMB VI and AMB IV) and the more recent level (AMB II). At PAP, the separation also occurs in the transition between the two oldest levels (PAP 5’25-6’25 and PAP 4’75-5’25) and the most recent one (PAP 4’00-4’75). These disruptions are, at both sites, coincident with the transition from the traditionally-defined Upper Solutrean to Solutreo-Gravettian (or Evolved Upper Solutrean) phases, which is thought to have occurred, at these sites, at around c. 20.5 ka cal BP [@munoz_ibanez_puntas_2000; @fullola_industrias_1979; @ripoll_lopez_cueva_1988]. Thus, previous differences detected between those phases, based on typological grounds and pointing to a decline in the use of flat-retouched points and the increasing presence of backed shouldered points [@aura_solutrenses_1989; @tiffagom_pierre_2006], seem to be corroborated by significant changes in the technological substrate. The most important changes regarding this transition comprise an increasing use of reduction sequences oriented to the production of highly elongated blanks, using opposed bidirectional methods on prismatically-organized cores. The absence of this tendency at VB [see e.g., @cascalheira_chronological_2015; @cascalheira_influencia_2013] might result from a lack of effective occupation at the site during the later parts of the LGM [which is rather probable given that the latest radiocarbon date for the solutrean occupations at the site is c. 21 ka cal BP - @cascalheira_chronological_2015; @cascalheira_vale_2012], but also by technological drift occuring during this period across Iberia, as suggested by the fact that a Solutreo-Gravettian phase is, so far, not clearly recognized in the Atlantic facade of Iberia.
